\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2019-07-17 popl-2019-journal-v/paper.tex
[POPLjv] Update references

To enforce these properties, target language security primitives are needed that can be used to prevent untrusted code from misbehaving without imposing too much overhead on well-behaved code.
The virtual-memory based security primitives on commodity processors do not seem sufficiently fine-grained to efficiently support this.
More suitable security primitives are offered by a type of CPUs known as capability machines \citep{levy_capability-based_1984,watson_cheri:_2015}.
These processors use tagged memory to enforce a strict separation between integers and {\itshape capabilities}: pointers that carry authority.
Capabilities come in different flavours.
Memory capabilities allow reading from and writing to a block of memory.
Additionally, capability machines offer some form of {\itshape object capabilities} that represent low-level encapsulated closures, i.e. a piece of code coupled with private state that it gains access to upon invocation.
The concrete mechanics of object capabilities vary between different capability machines.
% KJAA: This 'sealed' and 'common seal' business is unclear to me.
For example, on a recent capability machine called CHERI they take the form of pairs of capabilities that represent the code and data parts of the closure.
Each of the two capabilities are sealed with a common seal which make them opaque.
The hardware transparently unseals the pair upon invocation~\citep{watson_capability_2015,watson_fast_2016}.

@@

\begin{jversion}
  This text is an extended version of a paper that was presented at POPL 2019~\citep{skorstengaard_stktokens:_2019}.
  Compared to the earlier text, this version adds X and Y and Z.
  \todo{new version of TR?}
  This paper is accompanied by a technical report~\citep{technical_report_popl} with the elided details and proof.
\end{jversion}

@@

\section{A Capability Machine with Sealing and Linear Capabilities}
\label{sec:cap-mach-w-seal-and-lin}
% Brief intro to section and this cap machine:
% Cap machine inspired by CHERI, but with linear capabilities
In this section, we introduce a simple but representative capability machine with linear capabilities, that we call \trgcm{} (Linear Capability Machine).
\trgcm{} is mainly inspired by CHERI~\citep{watson_cheri:_2015} with linear capabilities as the main addition.
For simplicity, \trgcm{} assumes an infinite address space and unbounded integers.
% \lau{Do we have a good reason to assume an infinite address space for this work?
%   We do not model malloc, so we do not need to worry about what happens when malloc runs out of memory.}

@@

On \trgcm{} we have CHERI-like sealed capabilities to achieve this~\cite{watson_cheri:_2015,watson_fast_2016}.

@@

% Programs
We can now define the notion of a program as well as a context.
\begin{definition}[Programs and Contexts]
  \label{def:program-and-context}
  A \emph{program} is a component $(\comp_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})$ with an empty import list.
  A \emph{context} for a component $\comp$ is a component $\comp'$ such that $\comp \bowtie \comp'$ is a program.
\end{definition}
% Getting an executable from a component - postpone this to later, but mention:
% write xor exec
How a program is initialised to create an executable configuration, is discussed in Section~\ref{sec:form-secur-with}.
% We will, however, at this point mention that all programs will be instrumented as Write-XOR-Execute,{Do we have a reference for this?} i.e.\ there will only be read-execute capabilities for the code and read-write capabilities for the data.
%Further, we will only consider a specific class of well-formed components, but we will get back to this.
Some simplifications have been made in this presentation of \trgcm{}. See \citet{technical_report} for details.
\begin{figure}[tb]
  \centering
  \[
    \begin{array}{rl c rl c rl}
    s &\in \Symbol & \phantom{make} &
    \mathrm{import} &\mathrel{::=} a \mapsfrom s & \phantom{make} &
    \mathrm{export} &\mathrel{::=} s \mapsto w\\
    \comp_0 & \multicolumn{7}{l}{\mathrel{::=} (\mscode,\msdata,\overline{\mathrm{import}},\overline{\mathrm{export}},\sigrets,\sigcloss,A_\linear)}\\
    \comp & \multicolumn{7}{l}{\mathrel{::=} \comp_0 \mid  (\comp_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})}
  \end{array}
\]
\vspace{0.5cm}
\begin{mathpar}
  \inference{
    \comp_0 = (\mscode[1], \msdata[1], \overline{\var{import}_1}, \overline{\var{export}_1}, \sigrets[1], \sigcloss[1],A_{\linear,1})\\
    \comp_0' = (\mscode[2], \msdata[2], \overline{\var{import}_2}, \overline{\var{export}_2}, \sigrets[2], \sigcloss[2],A_{\linear,2})\\
    \comp_0'' = (\mscode[3], \msdata[3], \overline{\var{import}_3}, \overline{\var{export}_3}, \sigrets[3], \sigcloss[3],A_{\linear,3})\\
    \mscode[3] = \mscode[1] \uplus \mscode[2] \\
    \msdata[3] = (\msdata[1] \uplus \msdata[2])[a \mapsto w \mid (a \mapsfrom s) \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}), (s \mapsto w) \in \overline{\var{export}}_3] \\
    \overline{\var{export}_3} = \overline{\var{export}_1} \cup \overline{\var{export}_2}&
    \overline{\var{import}_3} = \{ a \mapsfrom s \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}) \mid s \mapsto \_ \not\in \overline{\var{export}_3} \}\\
    \sigrets[3] = \sigrets[1] \uplus \sigrets[2] &
    \sigcloss[3] = \sigcloss[1] \uplus \sigcloss[2] &
    A_{\linear,3} = A_{\linear,1} \uplus A_{\linear,2}\\
    \dom(\mscode[3]) \mathrel{\#} \dom(\msdata[3]) & \sigrets[3] \mathrel{\#} \sigcloss[3]
  } {
    \comp_0'' = \comp_0 \bowtie \comp_0'
  }
  \and
  \inference{
    \comp_0'' = \comp_0 \bowtie \comp_0'
  }{
    (\comp_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = \comp_0 \bowtie (\comp_0',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = (\comp_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}) \bowtie \comp_0'
  }
  \end{mathpar}
% \begin{definition}[Plugging a program into a context]
%   When $\var{comp'}$ is a context for component $\comp$ and $\comp' \bowtie \comp \rightsquigarrow \Phi$, 
%   then we write $\plug{\comp'}{\comp}$ for the execution configuration $\Phi$.
% \end{definition}
  \caption{Components and linking of components.}
  \label{fig:target-component-and-linking}
\end{figure}

@@

Capability machines originate with \citet{dennis_programming_1966} and we refer to \citet{levy_capability-based_1984} and \citet{watson_cheri:_2015} for an overview of previous work.
The capability machine formalized in Section~\ref{sec:cap-mach-w-seal-and-lin} is modelled after CHERI~\citep{watson_cheri:_2015,woodruff_cheri_2014}.
This is a recent, relatively mature capability machine which combines capabilities with a virtual memory approach in the interest of backwards compatibility and gradual adoption.
For simplicity, we have omitted features of CHERI that were not needed for \stktokens{} (e.g.\ local capabilities, virtual memory).

@@

CheriBSD applies a similar approach with separate per-component stacks and a trusted stack manager on a capability machine~\cite{watson_cheri:_2015}.
The authors use local capabilities to prevent components from accidentally leaking their stack pointer to other components, but there is no actual capability revocation in play.
They do not provide many details on this mechanism and it is, for example, not clear if and how they intend to deal with higher-order interfaces (C function pointers) or stack references shared across component boundaries.

>>>>

To enforce these properties, target language security primitives are needed that can be used to prevent untrusted code from misbehaving without imposing too much overhead on well-behaved code.
The virtual-memory based security primitives on commodity processors do not seem sufficiently fine-grained to efficiently support this.
More suitable security primitives are offered by a type of CPUs known as capability machines \citep{levy_capability-based_1984,watson_cheri_2015}.
These processors use tagged memory to enforce a strict separation between integers and {\itshape capabilities}: pointers that carry authority.
Capabilities come in different flavours.
Memory capabilities allow reading from and writing to a block of memory.
Additionally, capability machines offer some form of {\itshape object capabilities} that represent low-level encapsulated closures, i.e. a piece of code coupled with private state that it gains access to upon invocation.
The concrete mechanics of object capabilities vary between different capability machines.
% KJAA: This 'sealed' and 'common seal' business is unclear to me.
For example, on a recent capability machine called CHERI they take the form of pairs of capabilities that represent the code and data parts of the closure.
Each of the two capabilities are sealed with a common seal which make them opaque.
The hardware transparently unseals the pair upon invocation~\citep{watson_capability_2015,watson_fast_2016}.

@@

\begin{jversion}
  This text is an extended version of a paper that was presented at POPL 2019~\citep{skorstengaard_stktokens_2019}.
  Compared to the earlier text, this version adds X and Y and Z.
  \todo{new version of TR?}
  This paper is accompanied by a technical report~\citep{technical_report_popl} with the elided details and proof.
\end{jversion}

@@

\section{A Capability Machine with Sealing and Linear Capabilities}
\label{sec:cap-mach-w-seal-and-lin}
% Brief intro to section and this cap machine:
% Cap machine inspired by CHERI, but with linear capabilities
In this section, we introduce a simple but representative capability machine with linear capabilities, that we call \trgcm{} (Linear Capability Machine).
\trgcm{} is mainly inspired by CHERI~\citep{watson_cheri_2015} with linear capabilities as the main addition.
For simplicity, \trgcm{} assumes an infinite address space and unbounded integers.
% \lau{Do we have a good reason to assume an infinite address space for this work?
%   We do not model malloc, so we do not need to worry about what happens when malloc runs out of memory.}

@@

On \trgcm{} we have CHERI-like sealed capabilities to achieve this~\cite{watson_cheri_2015,watson_fast_2016}.

@@

% Programs
We can now define the notion of a program as well as a context.
\begin{definition}[Programs and Contexts]
  \label{def:program-and-context}
  A \emph{program} is a component $(\comp_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})$ with an empty import list.
  A \emph{context} for a component $\comp$ is a component $\comp'$ such that $\comp \bowtie \comp'$ is a program.
\end{definition}
% Getting an executable from a component - postpone this to later, but mention:
% write xor exec
How a program is initialised to create an executable configuration, is discussed in Section~\ref{sec:form-secur-with}.
% We will, however, at this point mention that all programs will be instrumented as Write-XOR-Execute,{Do we have a reference for this?} i.e.\ there will only be read-execute capabilities for the code and read-write capabilities for the data.
%Further, we will only consider a specific class of well-formed components, but we will get back to this.
Some simplifications have been made in this presentation of \trgcm{}. See \citet{technical_report_popl} for details.
\begin{figure}[tb]
  \centering
  \[
    \begin{array}{rl c rl c rl}
    s &\in \Symbol & \phantom{make} &
    \mathrm{import} &\mathrel{::=} a \mapsfrom s & \phantom{make} &
    \mathrm{export} &\mathrel{::=} s \mapsto w\\
    \comp_0 & \multicolumn{7}{l}{\mathrel{::=} (\mscode,\msdata,\overline{\mathrm{import}},\overline{\mathrm{export}},\sigrets,\sigcloss,A_\linear)}\\
    \comp & \multicolumn{7}{l}{\mathrel{::=} \comp_0 \mid  (\comp_0,c_{\mathrm{main},c}, c_{\mathrm{main},d})}
  \end{array}
\]
\vspace{0.5cm}
\begin{mathpar}
  \inference{
    \comp_0 = (\mscode[1], \msdata[1], \overline{\var{import}_1}, \overline{\var{export}_1}, \sigrets[1], \sigcloss[1],A_{\linear,1})\\
    \comp_0' = (\mscode[2], \msdata[2], \overline{\var{import}_2}, \overline{\var{export}_2}, \sigrets[2], \sigcloss[2],A_{\linear,2})\\
    \comp_0'' = (\mscode[3], \msdata[3], \overline{\var{import}_3}, \overline{\var{export}_3}, \sigrets[3], \sigcloss[3],A_{\linear,3})\\
    \mscode[3] = \mscode[1] \uplus \mscode[2] \\
    \msdata[3] = (\msdata[1] \uplus \msdata[2])[a \mapsto w \mid (a \mapsfrom s) \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}), (s \mapsto w) \in \overline{\var{export}}_3] \\
    \overline{\var{export}_3} = \overline{\var{export}_1} \cup \overline{\var{export}_2}&
    \overline{\var{import}_3} = \{ a \mapsfrom s \in (\overline{\var{import}_1} \cup \overline{\var{import}_2}) \mid s \mapsto \_ \not\in \overline{\var{export}_3} \}\\
    \sigrets[3] = \sigrets[1] \uplus \sigrets[2] &
    \sigcloss[3] = \sigcloss[1] \uplus \sigcloss[2] &
    A_{\linear,3} = A_{\linear,1} \uplus A_{\linear,2}\\
    \dom(\mscode[3]) \mathrel{\#} \dom(\msdata[3]) & \sigrets[3] \mathrel{\#} \sigcloss[3]
  } {
    \comp_0'' = \comp_0 \bowtie \comp_0'
  }
  \and
  \inference{
    \comp_0'' = \comp_0 \bowtie \comp_0'
  }{
    (\comp_0'',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = \comp_0 \bowtie (\comp_0',c_{\mathrm{main},c}, c_{\mathrm{main},d}) = (\comp_0,c_{\mathrm{main},c}, c_{\mathrm{main},d}) \bowtie \comp_0'
  }
  \end{mathpar}
% \begin{definition}[Plugging a program into a context]
%   When $\var{comp'}$ is a context for component $\comp$ and $\comp' \bowtie \comp \rightsquigarrow \Phi$, 
%   then we write $\plug{\comp'}{\comp}$ for the execution configuration $\Phi$.
% \end{definition}
  \caption{Components and linking of components.}
  \label{fig:target-component-and-linking}
\end{figure}

@@

Capability machines originate with \citet{dennis_programming_1966} and we refer to \citet{levy_capability-based_1984} and \citet{watson_cheri_2015} for an overview of previous work.
The capability machine formalized in Section~\ref{sec:cap-mach-w-seal-and-lin} is modelled after CHERI~\citep{watson_cheri_2015,woodruff_cheri_2014}.
This is a recent, relatively mature capability machine which combines capabilities with a virtual memory approach in the interest of backwards compatibility and gradual adoption.
For simplicity, we have omitted features of CHERI that were not needed for \stktokens{} (e.g.\ local capabilities, virtual memory).

@@

CheriBSD applies a similar approach with separate per-component stacks and a trusted stack manager on a capability machine~\cite{watson_cheri_2015}.
The authors use local capabilities to prevent components from accidentally leaking their stack pointer to other components, but there is no actual capability revocation in play.
They do not provide many details on this mechanism and it is, for example, not clear if and how they intend to deal with higher-order interfaces (C function pointers) or stack references shared across component boundaries.
